Expanding the Power of Lens Synthesis With Annotations

### 2019-02-12 papers/BXProposal/main.tex
minor changes before pulling bcp

\begin{abstract}
  Recent work has shown how to synthesize a class of string bijections from
  regular expression specifications. While these bijective string
  transformations can be useful for synchronizing data formats, it may still be
  useful to keep interesting formats synchronized when they are not in bijective
  correspondence. In this talk, we aim to show how we can expand the class of
  synthesized transformations beyond bijections, as long as users are willing to
  provide annotations that help identify a ``bijective core.''
\end{abstract}
\vskip 32pt

@@

Bidirectional languages are often difficult to program in. The can use
unfamiliar paradigms, and their type systems can sometimes be complex, making
fixing type errors oftentimes nontrivial. This is evident in the language
Boomerang~\cite{boomerang}, a DSL for writing lenses~\cite{Focal2005-long2}
on strings. In Boomerang, users can write bidirectional programs between data
formats: programs are lenses and types are regular expressions representing the
formats the lenses operate on. Instead of requiring users to manually write
bidirectional transformations, we find it preferable to synthesize bidirectional
transformations from their types.

@@

Bijections are an important class of bidirectional transformations, but many
useful transformations are not completely bijective. Boomerang permits defining
non-bijective transformations, such as lenses and quotient lenses, though
Optician cannot synthesize them. Lenses are bidirectional transformations that
may lose information when going from one side to the other, while obeying
certain roundtripping properties. Quotient lenses~\cite{quotientlenses} obey
similar properties to lenses, but also permit treating certain pieces of
information as irrelevant -- normalizing them to a canonical representative.

@@

At its base, this surrounding can be done solely with Optician and Boomerang,
but doing so requires repeated code. For example, consider the example of
converting an index of musical composers (with names, birth and death dates, and
nationalities) to a simple index of musical composers (with just names and
nationalities), but with irrelevant whitespace. In this example, ``.''
represents concatenation and ``\lstinline{*}'' represents the Kleene star.

@@

Consider the left-hand format of \lstinline{full_composers}. We expand our
regular expression types with the new combinator \lstinline{skip}. Annotating a
regular expression to be skipped does exactly what \lstinline{full_projection}
does in the previous code -- it sends the minified format to the core synthesis
engine, then composes the generated lens with a lens that deletes the regular
expression from the data format.
%
\begin{lstlisting}
  let full_composers' : annotated_regex = (name . (skip dates) . nationality)*
\end{lstlisting}
%

@@

>>>>

\begin{abstract}
  \emph{AFM: I will remove this from the submission, but am including it in the
    document for people to review. It is due today (February 12)} Recent work has shown how to synthesize a class of string bijections from
  regular expression specifications. While these bijective string
  transformations can be useful for synchronizing data formats, it may still be
  useful to keep interesting formats synchronized when they are not in bijective
  correspondence. In this talk, we aim to show how we can expand the class of
  synthesized transformations beyond bijections, for example into classical
  lenses, quotient lenses, and even symmetric lenses, as long as users are
  willing to provide annotations that help identify a ``bijective core.''
\end{abstract}
\vskip 32pt

@@

Learning any new domain-specific language takes time and energy. Any support we
can provide to help users ``get over the hump'' of learning a new DSL is going
to increase the rate of adoption for such a DSL. Indeed, due to the need to
think in ``two directions'' at once, bidirectional languages are more difficult
to learn than usual. This difficulty is evident in the language
Boomerang~\cite{boomerang}, a DSL for writing bidirectional programs on strings.
In Boomerang, users can write bidirectional programs between data formats.
Boomerang programs encode lenses a class of bidirectional objects. Boomerang
expressions are bidirectional programs on strings, and Boomerang types are
regular expressions representing the formats the lenses operate on. Instead of
requiring users to manually write bidirectional transformations, we find it
preferable to synthesize these transformations from their types.

@@

Bijections are an important class of bidirectional transformations, but many
useful transformations are not completely bijective. Boomerang expressions are
not necessarily bijections -- they encode lenses~\cite{Focal2005-long2}, and
various lens extensions like quotient lenses~\cite{quotientlenses}. Lenses are
bidirectional transformations that may lose information when going from one side
to the other, while obeying certain roundtripping properties. Quotient lenses
obey similar properties to lenses, but also permit treating certain pieces of
information as irrelevant -- normalizing them to a canonical representative.

@@

At its base, surrounding bijections in this way can be done solely with Optician
and Boomerang, but doing so requires repeated code. For example, consider the
example of converting an index of musical composers (with names, birth and death
dates, and nationalities) to a simple index of musical composers (with just
names and nationalities), but with irrelevant whitespace. In this example, ``.''
represents concatenation and ``\lstinline{*}'' represents the Kleene star.

@@

Consider the left-hand format of \lstinline{full_composers}. We expand our
regular expression types with the new combinator \lstinline{skip}. Annotating a
regular expression to be skipped does exactly what \lstinline{full_projection}
does in the previous code -- it sends the format, after removing the projected
portions, to the core synthesis engine, then composes the generated lens with a
lens that deletes the regular expression from the data format.
%
\begin{lstlisting}
  let full_composers' : annotated_regex = (name . (skip dates) . nationality)*
\end{lstlisting}
%

@@

Furthermore, we think this approach could scale to other lens extensions, like
matching lenses and dictionary lenses, and other types of bidirectional
constructs beyond merely lenses. All that is needed for this approach is a
user-provided a reduction to a ``bijective core,'' and a means to synthesize
that core. This means that, if a bijective synthesis algorithm is discovered for
(for example) trees or graphs, then annotations can be used to synthesize more
complex bidirectional programs for those structures as well.
